---
title: "431 Class 03 Demonstration"
author: "Thomas E. Love"
date: "2021-08-31"
output: 
    html_document:
        code_folding: show
        number_sections: TRUE
        toc_float: TRUE
        toc: TRUE
        df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# Setup

We're using R Markdown to gather together into a single document the code we build, text commenting on and reacting to that code, and the output of the analyses we build.

## Load packages and set theme

```{r, message = FALSE}
library(janitor)
library(knitr)
library(magrittr)
library(tidyverse)

theme_set(theme_bw())
```

Loading packages in R is like opening up apps on your phone. We need to tell R that, in addition to the base functions available in the software, we also have other functions we want to use. 

- The most important package is actually a series of packages called the `tidyverse`, which we'll use in every R Markdown file we create this semester. The `tidyverse` includes several packages, all developed (in part) by Hadley Wickham, Chief Scientist at RStudio.
    - `dplyr` is the primary package we'll use for data wrangling, cleaning and transformation
    - `ggplot2` is the primary package we'll use for visualization
    - other members of the `tidyverse` include packages for importing data, working with factors, and many other common activities.
- The `janitor` package has some tools for examining, cleaning and tabulating data (including `tabyl()` and `clean_names()`) that we'll use regularly.
- The `knitr` package has a function called `kable()` that we'll sometimes use to neaten up results.
- The `magrittr` package includes a pipe function denoted `%$%` which enables us to use "pipes" to pull specific variables out of a data frame (tibble) even when the function we want to use isn't part of the `tidyverse`.
- It's helpful to load the `tidyverse` package last.

# Ingest the data to `quicksur_raw`

The data we'll use is from the Quick Survey we did in Class 02 (pdf is available from the Class 02 README) and which I've also done in 2014-2020, in various forms.

## Today's Questions of Interest

A. What is the distribution of pulse rates among students in 431 since 2014?
B. Does the distribution of student heights change materially over time?
C. Do taller people appear to have paid less for their most recent haircut?
D. Does the relationship between height and haircut price look the same within each sex?
E. For the 2021 students specifically, do students have a more substantial tobacco history if they prefer to speak English or a language other than English?

## Read in data from `.csv` file

```{r}
quicksur_raw <- read_csv("data/quick_survey_2021.csv",
                         show_col_types = FALSE) %>%
    clean_names()
```

## What do we have?

```{r}
quicksur_raw
```

```{r}
glimpse(quicksur_raw)
```

## Counting Categories

```{r}
quicksur_raw %>% count(glasses)
```

```{r}
quicksur_raw %>% 
    filter(year == "2021") %>%
    tabyl(favcolor)
```


## Summarizing Quantities

```{r}
summary(quicksur_raw$pulse)
```

# Manage the data into `qsdat`

## Variables we'll look at today

Based on Questions A-E, we need to include the following variables in our analytic data frame.

-   `student`: student identification (numerical code)
-   `year`: indicates year when survey was taken (August)
-   `english`: y = prefers to speak English, else n
-   `smoke`: 1 = never smoker, 2 = quit, 3 = current
-   `pulse`: pulse rate (beats per minute)
-   `height_in`: student's height (in inches)
-   `haircut`: price of student's last haircut (in \$)
-   `sex`: f (for female) or m (for male)

## Select our variables from `quicksur_raw`

```{r}
qsdat <- quicksur_raw %>%
    select(student, year, english, smoke, 
           pulse, height_in, haircut, sex)
```

```{r}
qsdat
```

## Initial Numeric Summaries

-   Is everything the "type" of variable it should be? Are we getting the summaries we expect?

```{r}
summary(qsdat)
```

- Categorical variables should list the categories, with associated counts. To accomplish this, the variable needs to be represented in R with a `factor`, rather than as a `character` or `numeric` variable.

## Change the categorical variables described with numbers to factors

We want the `year` and `smoke` information treated as categorical, rather than as quantitative.

```{r}
qsdat <- qsdat %>%
    mutate(year = as_factor(year),
           smoke = as_factor(smoke))
```

## Change each character variable to a factor instead

We want to be able to summarize `english` and `sex`. We can use a little coding trick to change all of the character variables to factors.

```{r}
qsdat <- qsdat %>%
    mutate(across(where(is_character), as_factor)) 
```

- The `across(where())` syntax tells R to change everything that gives a TRUE response to "is this a character variable?" into a factor variable.

## Recheck the summaries and do range checks

-   Do these summaries make sense?
-   Are the minimum and maximum values appropriate?
-   How much missingness are we to deal with?

```{r}
summary(qsdat)
```

- Are we getting counts for all variables that are categorical?
    - Do the category levels make sense?
- Are we getting means and medians for all variables that are quantities?
    - Do the minimum and maximum values make sense for each of these quantities?
- Which variables have missing data, as indicated by `NA's`?
- Just to fill in the gap left by the table above, we might ask how many students we had responding each year.

```{r}
qsdat %>% count(year)
```

# Question A (Pulse Rates of our Students)

## A first try at a histogram

-   What is the distribution of student `pulse` rates?

```{r}
ggplot(data = qsdat, aes(x = pulse)) +
    geom_histogram(fill = "green", col = "blue")
```

How might we describe this distribution?

- What is the center?
- How much of a range around that center do we see? How spread out are the data?
- What is the shape of this distribution?
    - Is it symmetric, or is it skewed to the left or to the right? 

## Fundamental Numerical Summaries

```{r}
summary(qsdat$pulse)
```

- How do the summary statistics help us describe the data?
- Do the values make sense to you?

```{r, message = FALSE}
qsdat %$% mosaic::favstats(pulse)
```

- How can we measure variation (spread) in our data?

## Improving our Histogram

```{r}
qsdat %>%
    filter(complete.cases(pulse)) %>%
    ggplot(data = ., aes(x = pulse)) +
    geom_histogram(fill = "seagreen", col = "white",
                   bins = 20) +
    labs(title = "Pulse Rates of Dr. Love's students",
         subtitle = "2014 - 2021",
         y = "Number of Students",
         x = "Pulse Rate (beats per minute)")
```

- How did we deal with missing data?
- How did we add axis labels and titles to the plot?
- What is the distinction between `fill` and `col` in the histogram?
- What's the best choice for number of bins in the histogram?

# Question B (Student Heights over Time)

-   Does the distribution of student heights change materially over time, in the period of 2014-2021?

## Five-Number Summary of Heights for Each Year

```{r}
qsdat %>%
    filter(complete.cases(height_in)) %>%
    group_by(year) %>%
    summarize(n = n(),
              min = min(height_in),
              q25 = quantile(height_in, 0.25),
              median = median(height_in), 
              q75 = quantile(height_in, 0.75),
              max = max(height_in))
```

- Key summaries based on percentiles / quantiles
    - minimum = 0th, maximum = 100th, median = 50th
    - quartiles (25th, 50th and 75th percentiles)
    - Range is maximum - minimum
    - IQR (inter-quartile range) is 75th - 25th percentile
- These summaries are generally more resistant to outliers than mean, standard deviation
- Form the elements of a boxplot (box-and-whisker plot)

## Comparison Boxplot of Heights by Year

```{r}
qsdat %>%
    filter(complete.cases(height_in)) %>%
    ggplot(data = ., aes(x = year, y = height_in)) +
    geom_boxplot() +
    labs(title = "Heights of Dr. Love's students, by year",
         subtitle = "2014 - 2021",
         x = "Year",
         y = "Height (in inches)")
```

- How did we deal with missing data here?
- Box covers the middle half of the data (25th and 75th percentiles)
- Solid line indicates median
- Whiskers extend from the quartiles to the most extreme values that are not judged by Tukey's "fences" method to be candidate outliers
    - Fences are drawn at 25th percentile - 1.5 IQR and 75th percentile + 1.5 IQR
- Are any values candidate outliers by this method? For which years?
- Was it important to change `year` to a factor instead of a numeric variable earlier?

## Adding a Violin Plot to the Boxplot

- When we'd like to better understand the shape of a distribution, we can amplify the boxplot.

```{r}
qsdat %>%
    filter(complete.cases(height_in)) %>%
    ggplot(data = ., aes(x = year, y = height_in)) +
    geom_violin() +
    geom_boxplot(aes(fill = year), width = 0.3) +
    guides(fill = "none") +
    scale_fill_viridis_d() +
    labs(title = "Heights of Dr. Love's students, by year",
         subtitle = "2014 - 2021",
         x = "Year",
         y = "Height (in inches)")
```


- How did we change the boxplot when we added the violin?
- What would happen if we added the boxplot first and the violin second?
- What does `guides(fill = "none")` do?
- What does `scale_fill_viridis_d()` do?

## Table of Means and Standard Deviations

```{r}
qsdat %>%
    filter(complete.cases(height_in)) %>%
    group_by(year) %>%
    summarize(n = n(),
              mean = mean(height_in),
              sd = sd(height_in))
```

## The Empirical Rule for Approximately Normal Distributions

- If the data followed a Normal (or Gaussian) distribution, then approximately 68% of heights would be within 1 SD of the mean, and approximately 95% of heights would be within 2 SD of the mean.

In 2021, we had 55 students whose `height_in` was available.

- Do those 55 students appear to be well modeled by a Normal distribution, based on the histogram below and the boxplot with violin we saw previously?

```{r}
qsdat %>%
    filter(complete.cases(height_in)) %>%
    filter(year == "2021") %>%
    ggplot(data = ., aes(x = height_in)) +
    geom_histogram(fill = "salmon", col = "white",
                   binwidth = 1) +
    labs(title = "Heights Dr. Love's students",
         subtitle = "2021 (n = 55 students with height data)",
         y = "Number of Students",
         x = "Height (inches)")
```

- Of those, how many were within 1 SD of the mean, in other words, how many were in between 67.75 - 4.13 = 63.62 inches and 67.75 + 4.13 = 71.88 inches?

```{r}
qsdat %>% filter(complete.cases(height_in)) %>%
    filter(year == "2021") %>%
    count(height_in >= 63.62 & height_in <= 71.88)

34/(34+21)
```

- How many of the 55 `height_in` values gathered in 2021 were between 67.75 - (2)4.13 = 59.49 and 67.75 + 2(4.13) = 76.01?

```{r}
qsdat %>% filter(complete.cases(height_in)) %>%
    filter(year == "2021") %>%
    count(height_in >= 59.49 & height_in <= 76.01)

54/(54+1)
```


# Question C (Heights and Haircut Prices)

-   Do taller people pay less for their haircuts?

## Converting inches to centimeters

```{r}
qsdat <- qsdat %>%
    mutate(height_cm = height_in * 2.54)
```

## Initial Numerical Summaries

```{r}
qsdat %>%
    filter(complete.cases(haircut, height_cm)) %>%
    summarize(n = n(), median(haircut), 
              median(height_cm), median(height_in))
```

## A First Scatterplot

- We'll include the straight line from a linear model, in red.

```{r}
qsdat %>% filter(complete.cases(height_cm, haircut)) %>%
    ggplot(., aes(x = height_cm, y = haircut)) +
    geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm", col = "red",
                formula = y ~ x, se = TRUE) +
    labs(x = "Height (in cm)",
         y = "Price of last haircut (in $)",
         title = "Do taller people pay less for haircuts?")
```

## What is the (Pearson) correlation of height and haircut price?

```{r}
qsdat %>% 
    select(height_in, height_cm, haircut) %>%
    filter(complete.cases(.)) %>%
    cor(.) %>%
    kable(digits = 2)
```

## What is the straight line regression model?

```{r}
m1 <- qsdat %>% 
    filter(complete.cases(haircut, height_cm)) %$%
    lm(haircut ~ height_cm)

m1

summary(m1)
```

## Compare the `lm` fit to a `loess` smooth curve?

```{r}
qsdat %>% filter(complete.cases(height_cm, haircut)) %>%
    ggplot(., aes(x = height_cm, y = haircut)) +
    geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm", col = "red",
                formula = y ~ x, se = FALSE) +
    geom_smooth(method = "loess", col = "blue",
                formula = y ~ x, se = FALSE) +
    labs(x = "Height (in cm)",
         y = "Price of last haircut (in $)",
         title = "Do taller people pay less for haircuts?")
```

- Does a linear model appear to fit these data well?

# Question D (Height, Haircut Price and Sex)

- Does the haircut-height relationship look the same within each sex?

```{r}
qsdat %>% tabyl(sex)
```

## When do we see the missing `sex` data?

```{r}
qsdat %>% tabyl(sex, year)
```

## Some Numerical Summaries

```{r}
qsdat %>%
    filter(complete.cases(sex, haircut, height_cm)) %>%
    group_by(sex) %>%
    summarize(n = n(), median(haircut), median(height_cm))
```

## A scatterplot, identifying the sex of each subject

```{r}
qsdat %>% filter(complete.cases(height_cm, haircut, sex)) %>%
    ggplot(., aes(x = height_cm, y = haircut, 
                  group = sex, col = sex)) +
    geom_point() + 
    geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
    labs(x = "Height (in cm)",
         y = "Price of last haircut (in $)",
         title = "Do taller people pay less for haircuts?",
         subtitle = "accounting for sex")
```

## An alternative approach to identifying `sex`, with `facet_wrap`

```{r}
qsdat %>% filter(complete.cases(height_cm, haircut, sex)) %>%
    ggplot(., aes(x = height_cm, y = haircut, col = sex)) +
    geom_point() + 
    geom_smooth(method = "lm", col = "red",
                formula = y ~ x, se = FALSE) +
    geom_smooth(method = "loess", col = "blue",
                formula = y ~ x, se = FALSE) +
    facet_wrap(~ sex) +
    labs(x = "Height (in cm)",
         y = "Price of last haircut (in $)",
         title = "Do taller people pay less for haircuts?",
         subtitle = "accounting for sex")
```

## A third scatterplot looking at the same question

```{r}
qsdat %>% filter(complete.cases(height_cm, haircut, sex)) %>%
    ggplot(., aes(x = height_cm, y = haircut, col = sex)) +
    geom_point() + 
    geom_smooth(method = "lm", col = "red",
                formula = y ~ x, se = FALSE) +
    geom_smooth(method = "loess", col = "blue",
                formula = y ~ x, se = FALSE) +
    facet_grid(sex ~ .) +
    guides(col = "none") +
    labs(x = "Height (in cm)",
         y = "Price of last haircut (in $)",
         title = "Do taller people pay less for haircuts?",
         subtitle = "accounting for sex")
```

# Question E (Tobacco and Language Preference in 2021)

-   Do students in the 2021 class have a more substantial history of tobacco use if they prefer to speak a language other than English?

```{r}
qsdat2021 <- qsdat %>% 
    filter(year == "2021") %>%
    select(student, year, english, smoke)
```

```{r}
summary(qsdat2021)
```

No missing data.

## Tabulating the categorical variables individually

```{r}
qsdat2021 %>% tabyl(english)

qsdat2021 %>% tabyl(smoke) %>% adorn_pct_formatting()
```

- What does `adorn_pct_formatting()` do?

## Creating a cross-classification table (2 x 3)

```{r}
qsdat2021 %>% tabyl(english, smoke)
```

## Recode the `smoke` levels to more meaningful names in `tobacco`

```{r}
qsdat2021 <- qsdat2021 %>% 
    mutate(tobacco = fct_recode(smoke, 
            "Never" = "1", "Quit" = "2", "Current" = "3"))
```

Check our work?

```{r}
qsdat2021 %>% count(smoke, tobacco)
```

- Everyone with `smoke` = 1 has `tobacco` as Never, etc.

## Restate the cross-tabulation 

Now we'll use this new variable, and this time, add row and column totals.

```{r}
qsdat2021 %>% tabyl(english, tobacco) %>% 
    adorn_totals(where = c("row", "col"))
```

Don't forget to close your file with the session information.

# Session Information

```{r}
sessionInfo()
```

