---
title: "YOUR NAME: 431 Class 04 In Class Activity"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    code_folding: show
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)
```

## Setup

```{r load_packages, message = FALSE, warning = FALSE}

library(tidyverse)

## always need tidyverse, can include other packages too
## which should always be inserted above the tidyverse
```

## Import the `midwest` data

The midwest data is already in R, so you don't need to import it. However we will pull it from the package to our environment with the below code.

```{r}
midwest <- midwest
```

# Task 1: Load and Explore the Data

```{r}

```

# Task 2: Look at Cuyahoga County

```{r}

```

# Task 3: Make a boxplot

```{r}

```


# Task 4: Make a scatterplot

```{r}

```

# Task 5: Add a Loess smooth

```{r}

```

# Session Information

Adding a `sessionInfo()` chunk at the end of your document helps with reproducibility.  Take a look to see what it produces, and include the command at the end of this Lab, please.

```{r}
sessionInfo()
```

