---
title: "431 Class 04"
author: "Wyatt P. Bensken"
date: "2021-09-02"
output:
  beamer_presentation:
    theme: "Rochester"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_height: 5
    fig_caption: false
---

```{r set-options, echo=FALSE, cache=FALSE, message = FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 55)
```

# Welcome

## Today's Agenda

- Review Box plots, Scatterplots, and Loess smooth curves
  - IMS Chapter 5
- Hands on with R to explore data
- Walk through two figures
- Question(s)

## Upcoming Due Dates

- **Lab 01 is due Monday, 2021-09-06, at 9PM**

- As always, see our course website for the most recent course updates
  - [thomaselove.github.io/431](https://thomaselove.github.io/431/)

## A Note

- It's okay if you don't feel completely comfortable with R and building the visualizations we'll work on today!

- The goal of today is just to get hands on with R and some data to start building those foundational coding skills

- Be patient with yourself as you learn and don't be afraid to ask questions

# Introduction

## Important Visualizations

### Box plots

- Summarizes a dataset with 5 statistics, while identifying outliers
  - Median, Interquartile Range (IQR), Range
  - Outliers are generally marked as a point and are generally 1.5x IQR

### Scatterplots

- Used to visualize two numerical variables
- Each point is an observation
- Useful in assessing relationship between variables, and the trend

### Loess smooth curves

- We can fit a Loess smooth curve to the data, which can help reveal trends in the data which are not well estimated with a straight line.

## The data we'll use

- The `ggplot2` package contains the `midwest` dataset

```{r, echo = FALSE}
ls(ggplot2::midwest)
```

## Working with the data

### The variables we are interested in

- In this in-class work we are interested in the following variables:
  - `percbelowpoverty`, the percent of people below the poverty line
  - `percollege`, the percent who are college educated
  - `county`, the county name
  
### The tasks we'll accomplish

1. Load and Explore the Data
2. Look at Cuyahoga county (where we are now)
3. Make a boxplot
4. Make a scatterplot
5. Add a Loess smooth to our scatterplot

## R Markdown (.Rmd) Template

- There is a .Rmd (R Markdown) template available on today's README and and the Data Downloads page

- This is a template which you should download and save somewhere on your computer for today's activity

  - **Please follow the instructions provided, specific to your operating system, to download the template**
  
- Note: After Lab 01, all of your Labs will be completed using R Markdown. We have provided templates for Lab 02 and Lab 03

# Task 1. Load and Explore the Data

## Task 1. Load and Explore the Data

### Task 1a. Load the data

- You'll first want to load the tidyverse package we'll need, by running the below code

```{r, message = FALSE}
library(tidyverse)
```

- Next we'll want to load the midwest data into our environment (this isn't necessary, but makes things a bit more intuitive)

```{r}
midwest <- ggplot2::midwest
```

### Task 1b. Learn about the dataset

- By running the below code, we can open up (in our help tab) the documentation for the data

```{r}
?midwest
```

# Task 2. Look at Cuyahoga County

## Task 2. Look at Cuyahoga County

- To look at Cuyahoga County we'll need to `filter()` our data to just our observation of interest
  - We'll also `select()` only those two variables we'd like to look at
  - This is a good use of the pipe `%>%`
  
- We know from data documentation that `county` is our variable name, and from looking at the data we can see that the county names are all capitalized.
  - It is important to remember that R is case sensitive!
  
```r
midwest %>%
  filter(county == "CUYAHOGA") %>%
  select(county, percbelowpoverty, percollege)
```

## Task 2. Look at Cuyahoga County

```{r, echo = FALSE}
midwest %>%
  filter(county == "CUYAHOGA") %>%
  select(county, percbelowpoverty, percollege)
```

# Task 3. Make a boxplot

## Task 3. Make a boxplot

- Our goal will be to make a boxplot, of `percbelowpoverty` which looks like this

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

## Task 3. Make a boxplot

- We'll work through the code step by step, but the complete code looks like this:

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

## Task 3. Make a boxplot

### Step 1

- First we'll use ggplot to set our dataset and aesthetics (abbreviated "aes")
  - This code won't run anything, until we add our the geom we would like

```r
ggplot(data = midwest, aes(x = percbelowpoverty))
```

## Task 3. Make a boxplot

### Step 2

- Now we can add (note that we use `+` here and not the pipe) that we would like the boxplot geom.

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot()
```

## Task 3. Make a boxplot

### Step 2

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot()
```

## Task 3. Make a boxplot

### Step 3

- Each geom has a number of options specific to that type of figure, here we'd like to color our outliers red.

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red")
```

## Task 3. Make a boxplot

### Step 3

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red")
```

## Task 3. Make a boxplot

### Step 4

- **In this course no figure is complete without appropriate axes labels and titles**
- We can add (again use `+`) these using the `labs()` statement where we have x, title, and subtitle
  - There are numerous other options such as y, subtitle, and caption, that are available but that we don't use here.

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the 
                      midwest package in ggplot2")
```

## Task 3. Make a boxplot

### Step 4

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the midwest package in ggplot2")
```

## Task 3. Make a boxplot

### Step 5

- I'd like to get rid of that odd gray background which is, somewhat annoyingly, the default
- We can do this using a theme - `theme_bw()`

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw()
```

## Task 3. Make a boxplot

### Step 5

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Task 3. Make a boxplot

### Step 6 - the final figure

- Finally, in this boxplot the y-axis text and tick marks are not informative or helpful, so we'd like to remove them
- The `theme()` command has a whole host of options, but here we'll just use 2.

```r
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

## Task 3. Make a boxplot

### Step 6 - the final figure

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty)) +
  geom_boxplot(outlier.color = "red") +
  labs(x = "Percent of people below the poverty line",
       title = "Boxplot of poverty in Midwest counties",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

# Task 4. Make a scatterplot

## Task 4. Make a scatterplot

- Now we'd like to make a scatterplot with `percbelowpoverty` on our x-axis and `percollege` on our y-xis

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Task 4. Make a scatterplot

### Step 1

- Each figure we'll make in R using the `ggplot2` package will share substantial syntax, including the first step
- Here, however we assign not just an x aesthetic but a y aesthetic as well

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
```

## Task 4. Make a scatterplot

### Step 2 

- In this example, we want to add a point geom, which makes a scatterplot when we have properly assigned our x and y variables in the aesthetic
- We can again take advantage of the options within our geom to make the points a specific color.

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue")
```

## Task 4. Make a scatterplot

### Step 2 

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue")
```

## Task 4. Make a scatterplot

### Step 3

- As in our boxplot, we **must** add appropriate axis legends

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty 
                  and college eductation",
       subtitle = "These data come from the 
                      midwest package in ggplot2")
```

## Task 4. Make a scatterplot

### Step 3 

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation",
       subtitle = "These data come from the midwest package in ggplot2")
```

## Task 4. Make a scatterplot

### Step 4 - the final figure

- Finally, we'd like to again use our `theme_bw()` to get rid of that gray background

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty 
                  and college eductation",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw()
```

## Task 4. Make a scatterplot

### Step 4 - the final figure

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

# Task 5. Add a Loess smooth

## Task 5. Add a Loess smooth

- We'd now like to add a Loess smooth curve to our scatterplot to examine what type of relationship is fit with a smooth line.

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", color = "red", formula = y ~ x) +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation, with a Loess smooth curve",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 1

- One of the most powerful parts of ggplot and R is the ability to layer geoms
- We'll want to make sure to specify that we want a Loess curve in our method

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x) +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty 
                  and college eductation",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 1

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x) +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 2

- We can easily change the color of our Loess curve, to better differentiate it from the points.

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x, 
                                color = "red") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty 
                  and college eductation",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 2

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x, color = "red") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 3

- Finally, we'll want to update our title to reflect the new figure

```r
ggplot(data = midwest, aes(x = percbelowpoverty, 
                           y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x, 
                                color = "red") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty 
                  and college eductation,
                  with a Loess smooth curve",
       subtitle = "These data come from the 
                      midwest package in ggplot2") +
  theme_bw()
```

## Task 5. Add a Loess smooth

### Step 3

```{r, echo = FALSE}
ggplot(data = midwest, aes(x = percbelowpoverty, y = percollege)) +
  geom_point(color = "blue") +
  geom_smooth(method = "loess", formula = y ~ x, color = "red") +
  labs(x = "Percent of residents below poverty line",
       y = "Percent who are college educated",
       title = "Relationship between poverty and college eductation, with a Loess smooth curve",
       subtitle = "These data come from the midwest package in ggplot2") +
  theme_bw()
```

## Knit the file

- At the top of the R Markdown you should see a small button that says "Knit". Click this.

  - This turns your R Markdown into an HTML file
  
  - Again, this will be how you will complete Lab 02 - Lab 07

# Questions and Discussion

